Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-28 paper.tex
Add comparison graph (slam dunk). Loosen TDR language.

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. macOS has a timeout watchdog that, interestingly, does not terminate a spinning kernel---allowing us to obtain these results. On Linux, behavior varies by GPU driver.} (TDR), crashing the program.
\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{graphics/dist.pdf}
  \caption{Execution times of \emph{Decoupled Lookback} on an FPG-lacking device, Apple M1 Max. The TDR threshold here is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.\label{fig:dist}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
    Avg Spins per workgroup    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
    Avg Lookback Length per workgroup& 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, 32-bit Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}

>>>>

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’ 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{Ours}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{table*}
  \centering
  \small
  \setlength{\tabcolsep}{1pt}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r r r}
    \toprule
    Metric & \!\!\!\!\!\!Intel HD620 & ARM Mali-G78 MP20 & Apple M3 & Apple M1 Max & Nvidia 2080 Super & AMD 7900 XT \\
    \midrule
    Avg \emph{Memcpy} (32-bit ele/sec $\cdot 10^9$)  & 1.50   & 2.48  & 10.25  & 38.62  & 54.39  & 72.28  \\
    Avg \emph{DF} (32-bit ele/sec $\cdot 10^9$)   & 1.49   & 2.79  & 10.22  & 38.84  & 53.29  & 94.60  \\
    Avg \emph{RTS} (32-bit ele/sec $\cdot 10^9$)      & 1.06   & 2.02  & 6.81   & 25.13  & 35.90  & 61.49  \\
    \emph{DF}/\emph{Memcpy}       & 99.1\%    & 112.8\%  & 99.8\%    & 100.6\%   & 98.0\%    & 130.9\%   \\
    \emph{DF}/\emph{RTS}          & 140.2\%   & 138.5\%  & 150.2\%   & 154.5\%   & 148.4\%   & 153.8\%   \\
    Avg Spins per workgroup    & 0.305    & 1.477    & 0.606    & 0.604    & 0.419    & 0.148    \\
    Avg Lookback Length per workgroup& 1.726    & 3.537    & 3.415    & 10.782   & 13.046   & 20.722   \\
    Avg Total Fallbacks Initiated      & 254.662   & 2504.704  & 365.546   & 990.962   & 86.400   & 9.308   \\
    Avg Total Successful Insertions    & 6.170     & 122.436   & 4.612     & 0.012    & 0.284    & 0.020    \\
    \bottomrule
  \end{tabular*}
  \caption{\emph{Decoupled Fallback} performance, Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements.\label{tab:results}}
\end{table*}
